---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: references.bib
csl: new-phytologist.csl



---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\raggedright

**Title:** Nuisance species compromise the carbon sequestration potential in an Eastern US temperate deciduous forest


**Authors (tentative / incomplete list, not in order):** 

Luca Morreale

Krystal Bagnaschi

Rachel Hoffman

Iris Kennedy

Erin MacMonigle

Caroline Troy

Cameron Dow

Jen Jordan

Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)

David Mitre

Jess Shue

William McShea

Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)






**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, Virginia 22630, USA

2. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama


*corresponding author: teixeirak@si.edu; +1 540 635 6546

**Running head:** <!-- ≤45 characters-->

for submission to Anthromes joint special collection in New Phytologist/ Plants, People, Planet: https://www.newphytologist.org/news/view/332 

<!--JOURNAL

Target journal: 
Instructions for authors: 
article type:  -->


```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
# DISPLAY ITEMS
## I do not see any guidance as to the number of display items allowed in MEE.

#Tables
## example = 1 

#Figures
map_and_photos = 1
photo_deer_exclosure = "1a"
photo_represenative_upland = "1b"
photo_graveyard = "1c"

biomass_trends = 2 #line plot showing biomass through time for entire plot and each of the 3 focal sections

loss_gain_net = 3 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard

mortality_by_genera = 4 # 4-panel figure showing biomass mortality by genera for for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard

regen_failure = 5 #some figure illustrating absence of small canopy trees/ shifting  size distributions



```

\newpage

# Abstract  {-#Abstract} 

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (only outside deer exclosure?), reduced AGB (only outside deer exclosure?)
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

**Keywords**:  

\newpage

# Introduction  

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below. 

## (paragraph on tree mortality from non-endemic pests and pathogens)


Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
These can reduce C and need to be considered in future climate change projections.

@fei_biomass_2019

Non-endemic pests and pathogens facilitate invasin of non-endemic plants [@guo_impacts_2023 and refs therein].

## (paragraph on recruitment failure because of deer and non-endemic plants)

- white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence

@mcgarvey_effects_2013
@holm_interactive_2013
@knauer_understory_2023 

The capacity to regenerate following disturbances, including ongoing gap formation through mortality of canopy trees, critically influences long-term forest dynamics.
Regeneration depends first upon seed production and then upon seedling recruitment, survival, and growth into trees.
When any one of these steps fails, the stage is set for disturbance to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to it's pre-disturbance state in the foreseeable future [@refs].
Global change pressures can set the stage for such critical transitions by gradually shifting baseline conditions, making post-disturbance recovery unlikely despite the persistence of mature trees [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @refs]. 
In the mid-Atlantic region of eastern North America, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of current canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019;@miller_overabundant_2023]. 
Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023;@gorchov_differential_2021].
Deer are the biggest problem [@gorchov_differential_2021].
The juveniles that are present tend to represent a different, more mesophytic set of species (*Acer spp.*, *Fagus grandifolia*) than currently dominate much of the region (*Quercus spp.*, *Carya spp.*, @miller_compounding_2019; @nowacki_climate_2015]--a dynamic driven by fire suppression and mesophication [@refs].


Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) C stocks have declined (-∆AGB)
(2) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
(3) recruitment (outside deer exclosure) and growth have not kept pace with tree mortality, implying that future mortality will result in substantial net biomass loss (regeneration debt)


# Materials and Methods 

## Study site

*(describe location, forest type, etc)*

In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
(1) low deer, low non-endemic insects & pathogens (Fig. `r photo_deer_exclosure`)- 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
(2) high deer, low non-endemic insects & pathogens (Fig. `r photo_represenative_upland`) - # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens; 
(3) high deer, high non-endemic insects & pathogens (Fig. `r photo_graveyard`) - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens. 
These areas were delineated as follows.
Upland forest was defined according to topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
Vulnerability to non-endemic insects & pathogens was defined based on the abundance of canopy species affected by non-endemic pests and pathogens at the time of plot establishment (2008).


![**Figure `r map_and_photos`. Photos within the SCBI ForestGEO plot: (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens ** *(add a map of the plot)* All photos taken September 2023 by K. Anderson-Teixeira.](display/photos.png)

## Data collection

*(describe main census & years occurred)*

*(describe mortality census)*

*(describe invasive plants survey)*

*(describe seedlings survey, if used)*

## Analyses

apply the metric of @miller_overabundant_2023 to determine imminent failure/ probable failure/ insecure/ secure?

# Results 

# Discussion 

@holtmann_carbon_2021

# Conclusions (optional)

# Acknowledgements {-#Acknowledgements}

- Funding: ForestGEO


# Conflict of Interest statement {-#Conflict}
The authors declare no conflict of interest.

# Authors' contributions {-#Authors}
*[Name of author 1] and [Name of author 2] conceived the ideas and designed methodology; [Name of author 1] and [Name of author 3] collected the data; [Name of author 2] and [Name of author 4] analysed the data; [Name of author 1] and [Name of author 4] led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.*

# Data availability {-#Data}


\newpage
# References {-#References}




