---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: references.bib
csl: new-phytologist.csl



---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\raggedright

**Title:** Nuisance species compromise the carbon sequestration potential in an Eastern US temperate deciduous forest


**Authors (tentative / incomplete list, not in order):** 



Krystal Bagnaschi

Rachel Hoffman

Iris Kennedy

Erin MacMonigle

Caroline Troy

Cameron Dow

Jen Jordan

Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)

David Mitre

Jess Shue

Luca Morreale

William McShea

Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)






**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, Virginia 22630, USA

2. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama


*corresponding author: teixeirak@si.edu; +1 540 635 6546

**Running head:** <!-- ≤45 characters-->

for submission to Anthromes joint special collection in New Phytologist/ Plants, People, Planet: https://www.newphytologist.org/news/view/332 

<!--JOURNAL

Target journal: 
Instructions for authors: 
article type:  -->


```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
# DISPLAY ITEMS
## I do not see any guidance as to the number of display items allowed in MEE.

#Tables
## example = 1 

#Figures
## example = 1

```

\newpage

# Abstract  {-#Abstract} 

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (outside deer exclosure), reduced AGB (outside deer exclosure)?
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

**Keywords**:  

\newpage

# Introduction  

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023; @refs], but with declining CO~2~ sequestration capacity. <!--I think this is true, but check/ find more refs.-->
However, ... [@wu_uncertainty_2023]
<!--Global C models generally agree that the biome's CO~2~ sequestration capacity is likely to decline [@ahlstrom_robustness_2012].-->
Yet, global C models predict an wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014], including the effects of nuisance species. 

Nuisance species...

- often non-endemic
- white-tailed deer is important endemic nuisance species, over-abundant because of human influence


Tree mortality ...

Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].

The capacity to regenerate following disturbances, including ongoing gap formation through mortality of canopy trees, critically influences long-term forest dynamics.
Regeneration depends first upon seed production and then upon seedling recruitment, survival, and growth into trees.
When any one of these steps fails, the stage is set for disturbance to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to it's pre-disturbance state in the foreseeable future [@refs].
Global change pressures can set the stage for such critical transitions by gradually shifting baseline conditions, making post-disturbance recovery unlikely despite the persistence of mature trees [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @refs]. 
In the mid-Atlantic region of eastern North America, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of current canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019;@miller_overabundant_2023]. 
Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023].
The juveniles that are present tend to represent a different, more mesophytic set of species (*Acer spp.*, *Fagus grandifolia*) than currently dominate much of the region (*Quercus spp.*, *Carya spp.*, @miller_compounding_2019; @nowacki_climate_2015]--a dynamic driven by fire suppression and mesophication [@refs].

Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (or at least ∆AGB)
(3) C sequestration (∆AGB) has declined, and will decline further if current trends continue


# Materials and Methods 


# Results 

# Discussion 

# Conclusions (optional)

# Acknowledgements {-#Acknowledgements}

- Funding: ForestGEO


# Conflict of Interest statement {-#Conflict}
The authors declare no conflict of interest.

# Authors' contributions {-#Authors}
*[Name of author 1] and [Name of author 2] conceived the ideas and designed methodology; [Name of author 1] and [Name of author 3] collected the data; [Name of author 2] and [Name of author 4] analysed the data; [Name of author 1] and [Name of author 4] led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.*

# Data availability {-#Data}


\newpage
# References {-#References}




